---
title: "NYC Squirrels - DT Tidy Thursday"
author: "Patrick Anker"
date: "2019-01-30"
output: 
  prettydoc::html_pretty:
    theme: architect
    highlight: github
editor_options: 
  chunk_output_type: inline
---

# Goal

I chose the squirrel data because I had an interest in learning some of the mapping tools available in R, as well as in starting to get familiar with the `tidymodels` suite.

# Setup
First things first, load up and attach all of the packages needed in this document.
```{r packages}
library(tidyverse)
library(tidymodels)
library(magrittr)
library(skimr)
library(glue)

# Mapping packages
library(osmdata)
library(sf)

theme_set(theme_bw())
```

And load the squirrel dataset!
```{r}
squirrels <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-29/nyc_squirrels.csv")
```

# Exploration

Before looking at basic summaries of the data, let's insert the data dictionary for these data, supplied [here](https://github.com/rfordatascience/tidytuesday/blob/bf69a07555bd9925ccbf435bd2c211ecea0844b3/data/2019/2019-10-29/readme.md):

|variable                                   |class     |description |
|:---|:---|:-----------|
|long                                       |double    | Longitude|
|lat                                        |double    | Latitude|
|unique_squirrel_id                         |character | Identification tag for each squirrel sightings. The tag is comprised of "Hectare ID" + "Shift" + "Date" + "Hectare Squirrel Number." |
|hectare                                    |character | ID tag, which is derived from the hectare grid used to divide and count the park area. One axis that runs predominantly north-to-south is numerical (1-42), and the axis that runs predominantly east-to-west is roman characters (A-I).|
|shift                                      |character | Value is either "AM" or "PM," to communicate whether or not the sighting session occurred in the morning or late afternoon. |
|date                                       |double    | Concatenation of the sighting session day and month.|
|hectare_squirrel_number                    |double    | Number within the chronological sequence of squirrel sightings for a discrete sighting session.|
|age                                        |character | Value is either "Adult" or "Juvenile."|
|primary_fur_color                          |character | Value is either "Gray," "Cinnamon" or "Black."|
|highlight_fur_color                        |character | Discrete value or string values comprised of "Gray," "Cinnamon" or "Black."|
|combination_of_primary_and_highlight_color |character | A combination of the previous two columns; this column gives the total permutations of primary and highlight colors observed.|
|color_notes                                |character | Sighters occasionally added commentary on the squirrel fur conditions. These notes are provided here.|
|location                                   |character | Value is either "Ground Plane" or "Above Ground." Sighters were instructed to indicate the location of where the squirrel was when first sighted.|
|above_ground_sighter_measurement           |character | For squirrel sightings on the ground plane, fields were populated with a value of “FALSE.”|
|specific_location                          |character | Sighters occasionally added commentary on the squirrel location. These notes are provided here.|
|running                                    |logical   | Squirrel was seen running.|
|chasing                                    |logical   | Squirrel was seen chasing.|
|climbing                                   |logical   |  Squirrel was seen climbing.|
|eating                                     |logical   |  Squirrel was seen eating. |
|foraging                                   |logical   | Squirrel was seen foraging.|
|other_activities                           |character | Other activities   |
|kuks                                       |logical   | Squirrel was heard kukking, a chirpy vocal communication used for a variety of reasons.|
|quaas                                      |logical   | Squirrel was heard quaaing, an elongated vocal communication which can indicate the presence of a ground predator such as a dog.|
|moans                                      |logical   | Squirrel was heard moaning, a high-pitched vocal communication which can indicate the presence of an air predator such as a hawk.|
|tail_flags                                 |logical   | Squirrel was seen flagging its tail. Flagging is a whipping motion used to exaggerate squirrel's size and confuse rivals or predators. Looks as if the squirrel is scribbling with tail into the air.|
|tail_twitches                              |logical   | Squirrel was seen flagging its tail. Flagging is a whipping motion used to exaggerate squirrel's size and confuse rivals or predators. Looks as if the squirrel is scribbling with tail into the air.|
|approaches                                 |logical   | Squirrel was seen approaching human, seeking food.|
|indifferent                                |logical   | Squirrel was indifferent to human presence.|
|runs_from                                  |logical   | Squirrel was seen running from humans, seeing them as a threat.|
|other_interactions                         |character | Sighter notes on other types of interactions between squirrels and humans.|
|lat_long                                   |character | Combined lat long|
|zip_codes                                  |double    | zip codes|
|community_districts                        |double    | Community districts|
|borough_boundaries                         |double    | Borough boundaries|
|city_council_districts                     |double    | City council districts|
|police_precincts                           |double    | Police precincts |

Simple summmary:

```{r summary}
skim(squirrels)
```

Observations:
* Looks like the date is in milliseoncds from UNIX epoch, so that will need to be fixed.
* ZIP codes are completely missing, but that's fine since we have lat/long

Since we have longitude and latitude information, where are these observations?

```{r}
(popo <- squirrels %$% 
  table(police_precincts))

squirrels %$% 
  table(city_council_districts)
```

The vast majority of these observations are from Precinct `r names(popo)[which.max(popo)]`, so these are probably isolated to a single park. Specifically, according to a [CityLab article](https://www.citylab.com/life/2019/06/squirrel-census-results-population-central-park-nyc/592162/) about this dataset, this is wholly within Central Park.

Let's make a scatterplot of the observations.
```{r}
ggplot(squirrels) + 
  geom_point(aes(long, lat))
```

Each observation was recorded within a hectare block, given by the `hectare` variable.

```{r, echo = FALSE}
northsouth <- gsub("[[:alpha:]]$", "", squirrels$hectare) %>% 
  unique() %>% 
  as.integer()
```

The codes have two digits and one character, so I'm guessing the digits correspond to the north-south part of the grid since it has `r max(northsouth)` variations. I'll split the hectare codes into a grid.
```{r}
squirrels_work <- squirrels %>% 
  separate(hectare, c("hec_ns", "hec_ew"), sep = 2, remove = FALSE)

squirrels_hectare <- squirrels_work %>% 
  group_by(hectare) %>% 
  summarise(avg_long = mean(long), avg_lat = mean(lat), count = n())

squirrels_hectare %>% 
  ggplot() +
  geom_point(aes(avg_long, avg_lat, size = count)) +
  coord_fixed()
```

There are a few clear clusters of squirrel spotting, which look to correspond with human activity (note the clusters near clear entrances). Let's underlay a map of central park.

```{r}
bbox <- c(-73.983, 40.76429, -73.9474, 40.8)

central_park_query <- opq(bbox = bbox) %>% 
  add_osm_feature("leisure", "park") %>% 
  osmdata_sf()

central_park_poly <- central_park_query$osm_polygons %>% 
  filter(grepl("central", name, ignore.case = TRUE))

lake_query <- opq(bbox = bbox) %>% 
  add_osm_feature("natural", "water") %>% 
  osmdata_sf()
```

```{r}
squirrels_hectare %>% 
  ggplot() +
  geom_sf(data = central_park_poly, fill = "antiquewhite") +
  geom_sf(data = lake_query$osm_polygons, fill = "aliceblue") +
  geom_point(aes(avg_long, avg_lat, size = count)) +
  coord_sf(xlim = c(bbox[1], bbox[3]), ylim = c(bbox[2], bbox[4]))
  #   theme(panel.background = element_rect("aliceblue"))
```

Note the cluster to the north of "The Lake" (the larger southern lake): this is "The Ramble", a hilly and tree-covered hiking zone that leads to Belvedere Castle.

So kinds of squirrels are there?
```{r}
table(squirrels$primary_fur_color)
```
We have predominantly grey squirrels, which makes sense. But where are these cinnamon and black squirrels?

```{r}
fur_colors <- c(Cinnamon = "#E3816D", Gray = "#9F9F9F", Black = "#1F1F1F")

squirrels_work %>% 
  filter(!is.na(primary_fur_color)) %>% 
  ggplot() +
  geom_sf(data = central_park_poly, fill = "antiquewhite") +
  geom_sf(data = lake_query$osm_polygons, fill = "aliceblue") +
  geom_point(aes(long, lat, color = primary_fur_color)) +
  scale_color_manual(values = fur_colors) +
  coord_sf(xlim = c(bbox[1], bbox[3]), ylim = c(bbox[2], bbox[4]))
```

It looks like the grey and cinnamon squirrels are fairly uniformly distributed, but the black squirrels are found in little clusters! From here on in, I will create an azimuthal projection of Central Park and rotate a bit to "New York north" so that we can examine distributions of squirrel features without worrying about a squished coordinate system.

```{r}
central_park_centroid <- central_park_poly %>% 
  filter(name == "Central Park") %>% 
  st_geometry() %>% 
  st_centroid()

central_park_rotated <- central_park_poly %>% 
  st_set_crs(value = glue("+proj=aeqd +ellps=sphere +lat_0={}"))
```






